Testing the social, cognitive, and motor foundations of Paleolithic skill reproduction

### 2021-08-21 manuscript/manuscript.Rmd
updated the main text and supplementary tables.

```{r fig1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment. The demonstration flakes were made from the same basalt as used in the experiment with the same knapping technique."}
knitr::include_graphics(here::here("figure/Figure1.jpg"))
```

@@

Each participant was provided with nine nodules for use over the 2-hour experiment. We made these nodules from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This basalt has not been mechanically compared [e.g., rebound hardness, @braun2009] to East African basalts. Still, the basalt appears qualitatively like finer-grained examples at Gona (D. Stout, pers. obs.). Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that we could discriminate new fracture surfaces from those created during production (see supplementary Table 3). Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = 0.6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true for cores provided to participants across the two experimental conditions (Instructed vs. Uninstructed mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p= 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores used across participants and conditions could still emerge because of selection bias.

@@

```{r fig4, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects observing expert knapper in the untaught condition."}
knitr::include_graphics(here::here("figure/Figure4.jpg"))
```

@@

Continuous predictors were centered such that zero represents the sample average, and units are standard deviations. The full model was fitted with the lm function in R 3.2.3, and the *glmulti* package (Bartoń 2015) was used for multi-modal selection and model comparison.

@@

The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set (**Table** \@ref(tab:tab2)). Factor 1 tracks flake size with higher scores indicating larger flakes since all 11 measures load positively on this factor. Factor 2's loadings track the increasing relationship between thickness, length, and flake width. As factor 2 scores increase, flakes get thicker, longer, and narrow, resembling irregular splinters. Factor 3 tracks the relationship between flake proximal and distal width relative to thickness. As factor 3 scores go up, flakes get thinner and narrower at the distal ends and wider at the base. Factor 3 therefore tracks flakes with a typical shape having a thin cross-section, wider base, and narrower tip. We used these three flake shape coordinates to approximate DP size and shape in the project's flake performance factor analysis.
```{r tab2, echo=FALSE, message=FALSE}
library(tidyverse)
df2<-read.csv(here::here("data/tables/table2.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df2, booktabs = T, caption = "Lithic size/shape PCA factor loadings.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

To better understand the relationship between our various lithic performance measurements and to reduce data dimensionality, we conducted a second principal component analysis examining the study's six lithic performance measures (count of large pieces [\>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors). All of these measures were summarized for each core and unique factor scores were calculated from these core-specific measures. Bartlett's Test of Sphericity was significant ($\chi ^2$ (6) =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.

The analysis yielded two factors explaining a total of 56% of the variance for the entire set of variables (**Table** \@ref(tab:tab3)). Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and on core delta mass. Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation). High scores on Quality thus reflect production of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.
```{r tab3, echo=FALSE, message=FALSE}
library(tidyverse)
df3<-read.csv(here::here("data/tables/table3.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df3, booktabs = T, caption = "Overview of principal components results on the six flake performance measures. DP = detached piece.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

These two factors address flaking performance at the level of individual cores, however we were also interested in the overall productivity/rate of work of each participant over the entire two hours. For example, looking at a knappers average Quality and Quantity factor scores would not differentiate between a participant who spent the entire time exhaustively reducing one core vs. another participant who did the same to all nine of their allotted cores in the same time. To capture this aspect of variation we calculated a simple Total Productivity metric as the sum of all mass a participant removed from cores during the experiment.

@@

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (**Figure** \@ref(fig:fig5) and **Figure** \@ref(fig:fig6)). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (**Figure** \@ref(fig:fig5)a), although this relationship is twice as strong in the trained (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to untrained (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the untrained group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the Trained group (**Figure** \@ref(fig:fig5)b). A qualitatively similar trend with respect to Quality vs. Quantity (**Figure** \@ref(fig:fig5)c) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

@@

Thus, it appears that Trained participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality, whereas Untrained participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see variance Table and Figure) and generally increased productivity at the expense of Quality. Experimental artifacts illustrating these trade-offs are presented in **Figure???**.

@@

Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the trained and untrained groups. Our expert demonstrator/instructor is included as a performance benchmark.

**Table** \@ref(tab:tab4) summarizes the results of ANOVA tests group level difference in central tendency on various performance measures. We found no significant differences between the trained and untrained groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14), had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26), and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)). All three of these results show either medium or large effect sizes. In all three comparisons, the trained group's data distributions tended towards the expert sample although they were not significantly different from the untrained group (**Figure** \@ref(fig:fig8)). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values, but with a very low (\<0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance whereas novices in experimental groups produced pieces of similar mean size and shape as those of the expert trainer.
```{r tab4, echo=FALSE, message=FALSE}
library(tidyverse)
df4<-read.csv(here::here("data/tables/table4.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df4, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and with the expert instructor. All comparisons are three-way except for the two skill PC factors and the total mass removed, which are two-way comparisons between the subject population as these factors were not applied to the expert data sample. Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size). * = non-parametric permuted p-values to account for unequal sample variances. SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

```{r fig8, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of untaught, taught, and expert cores."}
knitr::include_graphics(here::here("figure/Figure8.jpg"))
```

While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the trained group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the trained and untrained individuals using the F-test on either core-averaged or flake specific variances. **Table** \@ref(tab:tab5) and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and total amount of flaked mass). In most instances, variance in the untrained group exceeds that of trained individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the untrained group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
```{r tab5, echo=FALSE, message=FALSE}
library(tidyverse)
df5<-read.csv(here::here("data/tables/table5.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df5)[colnames(df5) %in% c("X95..CI")] <- c("95% CI")
kableExtra::kbl(df5, booktabs = T, caption = "F-test results comparing variances between trained and untrained subjects across 
our various flaking performance metrics.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

```{r fig9, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Variance comparisons between trained and untrained individuals across various flaking performance metrics. Note the lower variance of the trained group in all the plots."}
knitr::include_graphics(here::here("figure/Figure9.jpg"))
```

@@

In addition to comparing overall performance during the two hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period. For these analyses, we calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8 both equal 50% complete). These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons. Flaking outcomes were tracked using the two performance factors (Quality and Quantity). We added the nodule starting mass to track whether training/practice times impacted raw material selection.

**Table** \@ref(tab:tab6) shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)). This result demonstrated that flaking outcomes did not change dramatically across the study interval. This lack of significant learning effects is confirmed by an inspection of individual learning curves (Figure). The one significant main training effect related to core starting mass (with a strong effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As the smaller nodules in their allotment were depleted, participants were left to knap larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the untrained group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the untrained group (Figure???).
```{r tab6, echo=FALSE, message=FALSE}
library(tidyverse)
df6<-read.csv(here::here("data/tables/table6.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df6, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and relative core order. All comparisons are three-way. Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size). SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

One of the experiment's primary goals was to test if measures of individual perceptual-motor and cognitive variation predict success in stone flaking across different training conditions. To address this goal, we built three multivariate models examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality). These models enabled us to determine which of the psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.

We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point. Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations. Our two motor skill and strength measures (grip strength and Fitt's performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41). However, these two measures track complementary components of athleticism (strength vs. speed/accuracy tradeoffs) and so we decided to include both in the model selection process.

The full models were fitted with the lm function in R 3.2.3, and we used the Glmulti package's automated model selection algorithm to select the best performing model (lowest AICc score) (see methods for further details on the multimodal selection process). All three models follow the same complete model statement as follows:

@@

For our two per-core performance factors (Quantity and Quality) it is also relevant to consider how individual core features may have affected performance. We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models. We did, however, find that subjects selected progressively larger nodules throughout the experiment. It is thus important to understand whether nodule variability had any impact on our flaking results. Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, $R^2$ = 0.54) we included nodule mass as a covariate to control for any variance in flaking performance that may be driven by nodule differences.

@@

Our first model examined variance in overall flaking productivity measured by each subject's combined flaked mass (nodule starting mass-core final mass). This provides a basic measure of variation in individuals' success detaching pieces and reducing cores from a standardized (see Methods) raw material supply. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13). This model comprised the following statement with two main and three interaction effects:

@@

**Table** \@ref(tab:tab7) presents this model's coefficients and summary outputs, wherein baseline refers to the untrained condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for 1 standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers at least when efficiency and quality are not considered.
```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
knitr::include_graphics(here::here("figure/Figure12.jpg"))
```

Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (**Figure** \@ref(fig:fig13)). In each case, higher scores were associated with better performance in the uninstructed group but worse performance in the instructed group. Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition (Coolidge and Wynn 2005) and social learning (Morgan et al. 2015) in the acquisition of knapping skills. Negative effects in the trained group are unexpected but presumably reflect differences in learning strategies adopted under instruction.

@@

**Table** \@ref(tab:tab8) presents this model’s coefficients and summary outputs, following the same format as Table. 
```{r tab8, echo=FALSE, message=FALSE}
library(tidyverse)
df8<-read.csv(here::here("data/tables/table8.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df8)[colnames(df8) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
kableExtra::kbl(df8, booktabs = T, caption = "Summary and estimates for Quantity flaking model.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength. As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the trained group but higher or unchanged Quantity in the untrained group (**Figure** \@ref(fig:fig14)).

@@

Unlike Total Productivity, the effect of Grip Strength on per-core Quantity was mediated by an interaction with learning condition (**Figure** \@ref(fig:fig15)). Thus, high Grip Strength enabled individuals in both groups to produce more total debitage, but only Instructed individuals translated Grip Strength into more intense reduction of individual cores, including not only delta mass, but also number and proportion of larger pieces.

@@

The Quality model did produce two statistically (p\< 0.05) significant interaction effects (RPM \* BEAST & Fitts \* n-back). However, these interactions had relatively small effects on Quality (\<0.5) and we believe that interpreting these results from our small, exploratory study would be inappropriate. **Figure** \@ref(fig:fig16) shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.

>>>>

```{r fig1, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects examining demonstration flakes prior to the experiment. The demonstration lakes were made from the same basalt as used in the experiment with the same knapping technique."}
knitr::include_graphics(here::here("figure/Figure1.jpg"))
```

@@

Each participant was provided with nine nodules for use over the 2-hour experiment. We made these nodules from larger chunks of a fine-grained basalt purchased from neolithics.com by fracturing them with a sledgehammer. This basalt has not been mechanically compared [e.g., rebound hardness, @braun2009] to East African basalts. Still, the basalt appears qualitatively like finer-grained examples at Gona (D. Stout, pers. obs.). Spalling produced irregular, angular chunks (**Figure** \@ref(fig:fig2)) for the experiment, weighing between 459g - 1876g (mean = 975g). All cores were weighed, measured (Length, Width, Thickness), and painted white so that we could discriminate new fracture surfaces from those created during production (see supplementary Table 1). Cores were sorted by shape and weight and then distributed evenly to each participant. As a result, there were no significant difference across participants in the mean weight (ANOVA, df = 22, F=0.3, p = 0.9; Levene test of homogeneity of variance = 1.04, df1=22, df2 = 184, p = 0.4) or shape (Length $\times$ Width/Thickness: ANOVA, df = 22, F=0.4, p = 0.9; Levene statistic = 0.6, df1=22, df2 = 184, p = 0.9) of cores provided. This was also true for cores provided to participants across the two experimental conditions (Instructed vs. Uninstructed mean weight = 1001g vs. 956g, t = 1.24, df = 205, p = 0.2, Levene's Test F = 0.6, p = 0.4; mean shape = 221.43 vs. 221.45, t = -0.003, df = 205, p = 0.9, Levene's Test F = 3.8, p= 0.05). Participants were, however, allowed to choose which cores to work on so that differences in the weight and shape of cores used across participants and conditions could still emerge because of selection bias.

@@

```{r fig4, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Subjects observing expert knapper in the uninstructed condition."}
knitr::include_graphics(here::here("figure/Figure4.jpg"))
```

@@

Continuous predictors were centered such that zero represents the sample average, and units are standard deviations. The full model was fitted with the *lm* function in R 4.0.3, and the *glmulti* package (Bartoń 2015) was used for multi-modal selection and model comparison.

@@

The analysis yielded three factors explaining a total of 90% of the variance for the entire 11 measurement variable set (**Table** \@ref(tab:tab2)). Factor 1 tracks size with higher scores indicating larger pieces since all 11 measures positively load this factor. Factor 2's loadings track increasing thickness and length but not width. As factor 2 scores increase, pieces thus get thicker, longer, and relatively narrower, resembling irregular splinters. Factor 3 tracks increasing proximal breadth and decreasing thickness. As factor 3 scores go up, pieces get thinner overall and broader at the base. Factor 3, therefore, tracks pieces with a more typical flat, flake-like morphology having a thin cross-section and tapering shape. We used these three shape coordinates to approximate DP size and shape in the project's flaking performance factor analysis.
```{r tab2, echo=FALSE, message=FALSE}
library(tidyverse)
df2<-read.csv(here::here("data/tables/table2.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df2, booktabs = T, caption = "Lithic size/shape PCA factor loadings.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)%>%
kableExtra::row_spec(1, bold = T)
```

@@

We conducted a second principal component analysis better to understand the relationship between our various lithic performance measurements and reduce data dimensionality. This analysis examined the study's six lithic performance measures (count of large pieces [\>40mm and 5g], mass of large pieces relative to total detached mass, core delta mass, and the three flake shape factors) (see supplementary Table 3 for details). These measures were summarized for each core, and individual factor scores were calculated from these core-specific measures. Bartlett's Test of Sphericity was significant ($\chi ^2(6)$  =3185, p \< .01) indicating that the set of variables are at least adequately related for factor analysis.

The analysis yielded two factors explaining 56% of the variance for the entire set of variables (**Table** \@ref(tab:tab3)). Factor 1 (hereafter "Quantity") explains 28.7% of the variance and tracks flaking quantity due to high positive loadings on large DP count and mass ratio and core delta mass. Performance factor 2 (hereafter "Quality") covers 27% of the sample variance and measures flaking quality as reflected in high positive loadings on Shape Factors 1 (size) and 3 (thin, "flake-like" shape) and a negative loading on Shape Factor 2 ("splinter-like" thickness and elongation). High scores on Quality thus reflect the making of larger, relatively thinner, and more typically flake-shaped vs. splinter-shaped DPs.
```{r tab3, echo=FALSE, message=FALSE}
library(tidyverse)
df3<-read.csv(here::here("data/tables/table3.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df3, booktabs = T, caption = "Overview of principal components results on the six flake performance measures. DP = detached piece.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)%>%
kableExtra::row_spec(1, bold = T)
```

These two factors address flaking performance at the level of individual cores. However, we were also interested in each participant's overall productivity/work rate over the entire two hours. For example, looking at a knappers average Quality and Quantity factor scores would not differentiate between a participant who spent the whole time exhaustively reducing one core vs. another participant who did the same to all nine of their allotted cores in the same time. We calculated a simple Total Productivity metric as the sum of all mass a participant removed from cores during the experiment to capture this aspect of variation.

@@

This approach also allowed us to compare the relationship between Total Productivity, Quantity, and Quality across our two experimental groups (**Figures** \@ref(fig:fig5) and \@ref(fig:fig6)). As might be expected, we found that per-core Quantity and Total Productivity are positively correlated in both groups (**Figure** \@ref(fig:fig5)a), although this relationship is twice as strong in the instructed  (F[1, 9] = 33, p \< 0.01, Adj. $R^2$ = 0.8) compared to uninstructed  (F[1, 8] = 8, p = 0.02, Adj. $R^2$ = 0.4) group. Interestingly, we also found evidence of a negative correlation between Total Productivity and Quality in the uninstructed group (F[1, 8] = 28, p = \<0.01, Adj. $R^2$ = 0.7), but no relation in the instructed group (**Figure** \@ref(fig:fig5)b). A qualitatively similar trend concerning Quality vs. Quantity (**Figure** \@ref(fig:fig5)c) did not achieve significance (F[1, 17] = 0.6, p = 0.2).

@@

Thus, it appears that instructed participants achieved higher Total Productivity by increasing average flaking Quantity across cores and without sacrificing Quality. In contrast, uninstructed participants found other ways to vary Total Productivity (e.g., number of cores knapped rather than Quantity per core, see **Table** \@ref(tab:tab5), **Figure** \@ref(fig:fig8) & \@ref(fig:fig9)) and generally increased productivity at the expense of Quality. Experimental artifacts illustrating these trade-offs are presented in **Figure** \@ref(fig:fig6).

@@

Here we compare our flaking outcomes (DP size/shape and flaking performance factors) between the instructed and uninstructed groups. We include the expert demonstrator/instructor as a performance benchmark.

**Table** \@ref(tab:tab4) summarizes the results of ANOVA tests group-level difference in central tendency on various performance measures. We found no significant differences between the instructed and uninstructed groups on our flaking Quantity and Quality factors. In contrast, three-way flake size and shape comparisons between our expert knapper and the two novice groups showed that the expert knapper made significantly more large flakes (effect size = 0.14). The expert also had a significantly higher core delta mass signal than either of the novice groups (effect size = 0.26) and left significantly smaller finished cores (effect size = 0.27) (**Figure** \@ref(fig:fig7)). All three of these results show either medium or large effect sizes. The instructed group's data distributions tended towards the expert sample in all three comparisons, although they did not differ significantly from the uninstructed group (**Figure** \@ref(fig:fig7)). We also observed a significant difference in shape factor 2 (splinters) driven by the expert's lower values but with a very small (\< 0.01) effect size. These results show that mean core reduction intensity and large flake production rates distinguish expert and novice performance. In contrast, novices in experimental groups produced similar mean detached piece sizes and shapes as the expert trainer. 
```{r tab4, echo=FALSE, message=FALSE}
library(tidyverse)
df4<-read.csv(here::here("data/tables/table4.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df4, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and with the expert instructor. All comparisons are three-way except for the two skill PC factors and the total mass removed, which are two-way comparisons between the subject population as these factors were not applied to the expert data sample. Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size). * = non-parametric permuted p-values to account for unequal sample variances. SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

```{r fig8, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Comparison of uninstructed, instructed, and expert cores."}
knitr::include_graphics(here::here("figure/Figure8.jpg"))
```

While we did not find significant differences in central tendency between our two experimental groups, results (**Figure** \@ref(fig:fig7)) did indicate lower variance in the instructed group. To test whether training reduced variability in performance outcomes between subjects, we compared variance metrics between the instructed and uninstructed individuals using the F-test on either core-averaged or flake-specific variances. **Table** \@ref(tab:tab5) and **Figure** \@ref(fig:fig9) present the results from these comparisons showing significant variance differences predominantly in flaking Quality, number of large DPs, core delta mass, and the total amount of flaked mass). In most instances, variance in the uninstructed group exceeds that of instructed individuals by 1.5 to 4.7 times. The most salient effect of instruction was thus not to shift mean performance but to reduce variability by eliminating the skew (generally toward poorer outcomes) seen in the uninstructed group (**Figure** \@ref(fig:fig9)), rather than to shift the mean.
```{r tab5, echo=FALSE, message=FALSE}
library(tidyverse)
df5<-read.csv(here::here("data/tables/table5.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df5)[colnames(df5) %in% c("X95..CI")] <- c("95% CI")
kableExtra::kbl(df5, booktabs = T, caption = "F-test results comparing variances between trained and untrained subjects across 
our various flaking performance metrics.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

```{r fig9, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Variance comparisons between instructed and uninstructed individuals across various flaking performance metrics. Note the lower variance of the trained group in all the plots."}
knitr::include_graphics(here::here("figure/Figure9.jpg"))
```

@@

In addition to comparing overall performance during the two-hour experiment, we also wanted to determine if groups or individuals differed in learning (i.e., performance change) over the period.  We calculated the learning stage as the ordinal number of each core out of the total number knapped by each subject (i.e., core 2 of 4 or 4 of 8, both equal 50% complete) for these analyses. These relative core use-order percentages were then binned into 20 percent brackets for core-order and group-level comparisons. We tracked flaking outcomes using the two performance factors (Quality and Quantity). We added the nodule starting mass to track whether training/practice times impacted raw material selection.

**Table** \@ref(tab:tab6) shows no significant training effects across the two performance measures either as grouped data or between individuals (**Figure** \@ref(fig:fig10) and **Figure** \@ref(fig:fig11)). This result demonstrated that flaking outcomes did not change dramatically across the study interval. An inspection of individual learning curves confirms this lack of significant learning effects (**Figure** \@ref(fig:fig11)). The main training effect was core starting mass (with a substantial effect size = 0.25). On average, core starting masses start low and increase, showing that participants selected smaller nodules first. As participants depleted the smaller nodules in their allotment, they knapped larger, less preferred nodules. This preference for smaller cores is somewhat less pronounced in the uninstructed group, as indicated by a small main effect of learning condition and generally higher starting nodule masses for the uninstructed group (**Figure** \@ref(fig:fig11)). 
```{r tab6, echo=FALSE, message=FALSE}
library(tidyverse)
df6<-read.csv(here::here("data/tables/table6.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df6, booktabs = T, caption = "Summary group-wise comparison statistics contrasting flake performance metrics by experiment condition and relative core order. All comparisons are three-way. Eta2 = ANOVA effect size (>0.1 = medium effect size, >0.2 = large effect size). SS = sum of squared differences, df = degrees of freedom, MS = mean squared difference.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

@@

One of the experiment's primary goals was to test if individual perceptual-motor and cognitive variation measures predict success in stone flaking across different training conditions. We built three multivariate models to address this goal examining the relations between training conditions, individual difference measures, and our three lithic performance measures (overall productivity and average per-core Quantity and Quality). These models enabled us to determine which psychometric and motor skill factors are better predictors of a participant's flaking performance in the study.

We considered all possible interactions between five individual difference measures, core size, training condition, and the three performance measures, with each subject providing one data point. Each model's continuous predictors (highest n-back level, Raven's Progressive Matrix score, BEAST score, starting nodule mass, Fitts score, and grip strength) were centered such that zero represents the sample average, and units are standard deviations. Our two motor skill and strength measures (grip strength and Fitt’s performance scores) are also strongly correlated (F [1,19] = 15, p \< 0.01, $R^2$ = 0.41). However, these two measures track complementary components of athleticism (strength vs. speed/accuracy trade-offs), so we decided to include both in the model selection process.

All three models follow the same complete model statement as follows:

@@

For our two per-core performance factors (Quantity and Quality), it is also relevant to consider how individual core features may have affected performance. We found no evidence of individual or group level practice effects over the two hours, so we did not include core order in the models. We did, however, find that subjects selected progressively larger nodules throughout the experiment. It is thus essential to understand whether nodule variability had any impact on our flaking results. Because starting nodule size (mass) and shape were strongly correlated (F [1,157] = 186, p \< 0.01, $R^2$ = 0.54), we included nodule mass as a covariate to control for any variance in flaking performance that nodule differences may drive.

@@

Our first model examined variance in overall flaking productivity measured by each subject's combined flaked mass (nodule starting mass-core final mass). This productivity metric provides a basic measure of variation in individuals' success detaching pieces and reducing cores from a standardized (see Methods) raw material supply. From the same candidate pool size of 55893 possible multivariate models, the best performing model returned an AICc value of -18 (Average AIC = -13). This model comprised the following statement with two main and three interaction effects:

@@

**Table** \@ref(tab:tab7) presents this model's coefficients and summary outputs, wherein baseline refers to the uninstructed condition with all continuous predictors at the sample average. The parameter estimates for the continuous predictors reflect the expected change in utility for one standard deviation change in the predictor variable. We found significant (p\< 0.05) and substantial (Standardized Estimate \>= 0.50, i.e. a 50% change in variable) main effects of Grip Strength, Visuospatial nBack, and BEAST. The main effect of Grip Strength (**Figure** \@ref(fig:fig12)), irrespective of learning condition, indicates the basic importance of strength in generating higher production rates among naive knappers, at least when we do not consider efficiency and quality.
```{r fig12, echo=FALSE, out.width = '100%', fig.align='center', fig.cap="Significant main effect of grip strength on overall flaking productivity."}
knitr::include_graphics(here::here("figure/Figure12.jpg"))
```

Effects of visuospatial working memory capacity and social information use are more complicated, as indicated by strong interactions with learning condition (**Figure** \@ref(fig:fig13)). In each case, higher scores were associated with better performance in the uninstructed group but *worse* performance in the instructed group. Positive effects in the uninstructed group were as expected, given the hypothesized importance of spatial cognition [@coolidge2005] and social learning [@morgan2015] in acquiring knapping skills. Negative effects in the instructed group are unexpected but presumably reflect differences in learning strategies adopted under instruction.

@@

**Table** \@ref(tab:tab8) presents this model’s coefficients and summary outputs, following the same format as **Table** \@ref(tab:tab6) and **Table** \@ref(tab:tab7).
```{r tab8, echo=FALSE, message=FALSE}
library(tidyverse)
df8<-read.csv(here::here("data/tables/table8.csv"), fileEncoding = 'UTF-8-BOM')
colnames(df8)[colnames(df8) %in% c("X95..CI", "t..12.")] <- c("95% CI", "t(12)")
kableExtra::kbl(df8, booktabs = T, caption = "Summary and estimates for Quantity flaking model.") %>%
  kableExtra::kable_styling(latex_options = c("striped", "HOLD_position"), font_size = 7)%>%
kableExtra::row_spec(0, bold = T)
```

The Quantity model roughly paralleled results for Total Production, yielding substantial and significant interactions between training condition, n-back level, BEAST scores, and grip strength. As with Total Production, higher visuospatial n-back levels and BEAST scores were associated with lower Quantity scores in the instructed group but higher or unchanged DP Quantity in the uninstructed group (**Figure** \@ref(fig:fig14)).

@@

Unlike Total Productivity, an interaction with learning condition mediated the effect of Grip Strength on per-core Quantity (**Figure** \@ref(fig:fig15)). Only instructed individuals translated Grip Strength into a more intense reduction of individual cores, including delta mass and the number and proportion of larger pieces. Thus, high Grip Strength enabled individuals in both groups to produce more debitage.

@@

The Quality model did produce two statistically (p\< 0.05) significant interaction effects (RPM \* BEAST & Fitts \* n-back). However, these interactions had relatively small effects on Quality (\<0.5), and we believe that interpreting these results from our small, exploratory study would be inappropriate. **Figure** \@ref(fig:fig16) shows the uneven distribution of data points for these interactions, which suggests vulnerability to leveraging effects of a small number of extreme value combinations.
